---
title: "Kevin Gerena Assignment 12 - Connectivity"
output: html_notebook
---


```{r}
require(tidyverse)
require(terra)
require(sf)
require(gdistance)
require(igraph)
require(fitdistrplus)
require(fdrtool)
require(sf)
require(AICcmodavg)
require(tidyterra)
require(dplyr)
require(raster)
require(landscapemetrics)
require(PerformanceAnalytics)
```


# Challenge 1 (5 points)

In the lab, we calculated Euclidean distance, least-cost distance, commute distance (based on circuit theory), and a randomized shortest path distance. Examine the correlation among these 4 metrics of effective distance. Which tend to be the longest, which tend to be the shortest, and why? In what situations might you expect to find strong differences between Euclidean distance and the other metrics of effective distance?

```{r}
#import data
# Read in the landcover raster
land = rast('https://github.com/ValenteJJ/SpatialEcology/raw/main/Week10/panther_landcover.tif')

# Bring in the public lands shapefile
public = st_read('/vsicurl/https://github.com/ValenteJJ/SpatialEcology/raw/main/Week12/panther_publicland.shp')

# Set the crs of the public lands shapefile
st_crs(public) = crs(land)

# And look at it
data.frame(public)

publicCentroids = st_centroid(public)

classification = read.table('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week12/resistance%20reclass.txt', header=T)
landCost = classify(land, classification[,c(1,3)])
landCost = raster(landCost)

#euclidean distance
geoDist = pointDistance(publicCentroids, lonlat=F)
geoDist = as.dist(geoDist)
geoDist
g1 = c(127755.36, 64612.56, 112427.46, 88783.37) #TOTACRES

#least-cost distance
landCond = transition(1/landCost, transitionFunction = mean, 8)
landCond = geoCorrection(landCond, type='c', multpl=F)
lcDist = costDistance(landCond, st_coordinates(publicCentroids))
lcDist
l1 = c(250607.85, 179140.02, 318972.02, 191449.65)

#commute distance
circuitDist = commuteDistance(landCond, st_coordinates(publicCentroids))
circuitDist
c1 = c(1337632.6, 1221718.7, 1675535.7, 1271736.3)

#randomized shortest path distance
rspDist1 = rSPDistance(landCond, from=st_coordinates(publicCentroids)[1,], to=st_coordinates(publicCentroids)[2,], theta=0.001)
rspDist2 = rSPDistance(landCond, from=st_coordinates(publicCentroids)[1,], to=st_coordinates(publicCentroids)[3,], theta=0.001)
rspDist3 = rSPDistance(landCond, from=st_coordinates(publicCentroids)[1,], to=st_coordinates(publicCentroids)[4,], theta=0.001)
rspDist4 = rSPDistance(landCond, from=st_coordinates(publicCentroids)[1,], to=st_coordinates(publicCentroids)[5,], theta=0.001)
r1 = rbind(rspDist1, rspDist2, rspDist3, rspDist4)

#examining correlation
Dist = data.frame(
  Euclidean = as.vector(g1),
  LeastCost = as.vector(l1),
  Circuit = as.vector(c1),
  Random = as.vector(r1))
Dist
chart.Correlation(Dist, histogram=T)
```

The randomized shortest distances are the longest, and the Euclidean distances are the shortest. This is expected because the Euclidean distance is a straight-line distance that does not include factors like fragmentation which may lead to underestimating the distance between locations in circumstances where factors such as that may be significant.

$\color{red}{\text{Good. +5}}$

# Challenge 2 (5 points)

In the lab we developed a 10% least-cost corridor between two protected areas. Identify what proportion of this corridor is comprised of each landcover type based on the land raster. To do this, make sure that both the land and leastCostCorridor10 rasters are spatRasters. Then mask land with leastCostCorridor10. The values in the resulting raster will be representative of the landcover values in the corridor. Summarise this information and refer back to the classification dataframe to link the landcover numbers with landcover types. What are the three most common landcover types within this corridor, and does this make sense given what you know about the resistance to movement within such landcover types?

```{r}
#mapping least cost paths
# Choose the extent for cropping
fpwrOssfExtent = extent(642000, 683000, 237000, 298000)

# Crop the landcover layer
landSub = crop(land, fpwrOssfExtent)

# Crop the cost layer and create a transition matrix
landCostSub = crop(landCost, fpwrOssfExtent)
landCondSub = transition(1/landCostSub, transitionFunction=mean, 8)
landCondSub = geoCorrection(landCondSub, type='c', multpl=F)

fpwrCost = accCost(landCondSub, st_coordinates(publicCentroids)[5,])
ossfCost = accCost(landCondSub, st_coordinates(publicCentroids)[3,])
plot(fpwrCost)
plot(ossfCost)

fpwrOssfLcp = shortestPath(landCond, st_coordinates(publicCentroids)[5,], st_coordinates(publicCentroids[3,]), output='SpatialLines')

plot(landCostSub, axes=F)
plot(st_geometry(public), add=T)
plot(st_geometry(publicCentroids), col='grey30', add=T)
lines(fpwrOssfLcp, col='red', lwd=3)

leastCostCorridor = overlay(fpwrCost, ossfCost, fun=function(x, y){return(x+y)})
plot(leastCostCorridor)

quant10 = quantile(leastCostCorridor, probs=0.1, na.rm=T)
quant10
leastCost10 = leastCostCorridor
values(leastCost10) = NA
leastCost10[leastCostCorridor < quant10] = 1
leastCost10 = rast(leastCost10)

plot(leastCost10, legend=F, axes=F)
points(publicCentroids, col='grey30')
lines(fpwrOssfLcp, col='red')

landMask = crop(land, leastCost10)
landMask = mask(landMask, leastCost10)
plot(landMask)

landMask = categories(landMask, value = classification[,c(1,3)])
landProp = lsm_c_pland(landMask, directions = 8)
landProp
classification
```

The three most common cover types are cypress swamp (22), cropland (14.3), and fresh water marsh (13.7). Cypress swamp is related to resource selection and cropland and fresh water marsh can also provide additional resources or terrain that facilitates travel.

$\color{red}{\text{I'm not exactly sure what went wrong here, but hardwood swamp should be in the top 3, not cropland.  Either way, good work. +4}}$


# Challenge 3 (5 points)

In the lab, we used the passage() function to conduct a randomized shortest path analysis mapping connectivity among two protected areas. We initially set theta to 0 thus resulting in a circuit theory based map of connectivity. Using the same passage() function, conduct the analysis again using values of theta = 0.0001 and theta = 0.001. Plot the resulting maps. What patterns do you see emerge as theta increases? Comment on the tradeoffs between biological realism and ease of conservation planning based on these three different maps of connectivity.

```{r}
#0 
passageMapT0 = passage(landCondSub, origin = st_coordinates(publicCentroids)[3,], goal = st_coordinates(publicCentroids)[5,], theta=0)

#0.0001
passageMapT1 = passage(landCondSub, origin = st_coordinates(publicCentroids)[3,], goal = st_coordinates(publicCentroids)[5,], theta=0.0001)

#0.001
passageMapT2 = passage(landCondSub, origin = st_coordinates(publicCentroids)[3,], goal = st_coordinates(publicCentroids)[5,], theta=0.001)

plot(passageMapT0)
plot(passageMapT1)
plot(passageMapT2)
```
As theta increases, the movement path becomes clearer and more readily applied to conservation. However, this comes at the cost of biological realism with lower theta values likely better reflecting reality or our lack of knowledge regarding contributing factors.

$\color{red}{\text{Good. +5}}$


# Challenge 4 (5 points)

In the latter part of the lab, we discussed calculating patch-based and landscape-based metrics of connectivity. Patch number 6 had the lowest degree (n = 2) while patch number 7 had one of the highest degrees (n = 12). First delete patch number 6 from the network, and then examine the impacts on the landscape level metrics integral index of connectivity (ICC) and probability of connectivity (PC). Now replace patch 6 back into the network and instead delete patch number 7. Again, examine how ICC and PC change. How does the loss of patch 6 compare to the loss of patch 7 in terms of its impacts on landscape-level connectivity?

```{r}
#patch-based networks
nodes = read.csv('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week12/kite_nodes.csv')
area = nodes$area
nodes

aObs = read.csv('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week12/kite_movement.csv')[,-1]
diag(aObs) = 0
aObs

coords = cbind(nodes$XCoord, nodes$YCoord)
distMat = pointDistance(coords, lonlat=F)
distMat = distMat / 1000 # in km
distMat

#dispersal kernels
linkLoc = which(aObs > 0, arr.ind=T)
withinDisp = cbind(distMat[linkLoc], aObs[linkLoc])

#Creating one distance observation for each individual
withinDisp = rep(withinDisp[,1], withinDisp[,2])

#fit model
dispExp = fitdist(data = withinDisp, distr='exp', method='mle')
plot(dispExp)
aObs #directed, weighted network

aMeanDist = mean(withinDisp)
aMeanDist #undirected, unweighted network based on mean dispersal distance

aMean = matrix(0, nrow=nrow(aObs), ncol=ncol(aObs))
aMean[distMat < aMeanDist] = 1
diag(aMean) = 0
aMean #transition matrix

#exponential distribution distance matrix
aProb = matrix(0, nrow=nrow(aObs), ncol=ncol(aObs))
alpha = 1/aMeanDist
aProb = exp(-alpha*distMat)
diag(aProb)=0
aProb

#igraph objects for each matrix
graphAmean = graph_from_adjacency_matrix(aMean, mode='undirected')
graphAprob = graph_from_adjacency_matrix(aProb, mode='undirected', weighted=T)
graphAobs = graph_from_adjacency_matrix(as.matrix(aObs), mode='directed', weighted=T)

plot(graphAmean, layout=coords, vertex.label=NA)
plot(graphAprob, layout=coords, edge.width=E(graphAprob)$weight, vertex.label=NA)
plot(graphAobs, layout=coords, vertex.label=NA)

```

### Patch 6 Removed
```{r}

nodes6 = read.csv('https://github.com/rytomey/WILD7970_Spatial/raw/main/W12_Connectivity/kite6.csv')
area6 = nodes6$area
nodes6

aObs6 = read.csv('https://github.com/rytomey/WILD7970_Spatial/raw/main/W12_Connectivity/kitemove6.csv')[,-1]
diag(aObs6) = 0
aObs6

coords6 = cbind(nodes6$XCoord, nodes6$YCoord)
distMat6 = pointDistance(coords6, lonlat=F)
distMat6 = distMat6 / 1000 # in km
distMat6

linkLoc6 = which(aObs6 > 0, arr.ind=T)
withinDisp6 = cbind(distMat6[linkLoc6], aObs6[linkLoc6])
withinDisp6 = rep(withinDisp6[,1], withinDisp6[,2])

dispExp6 = fitdist(data = withinDisp6, distr='exp', method='mle')
plot(dispExp6)
aObs6

aMeanDist6 = mean(withinDisp6)
aMeanDist6

aMean6 = matrix(0, nrow=nrow(aObs6), ncol=ncol(aObs6))
aMean6[distMat6 < aMeanDist6] = 1
diag(aMean6) = 0
aMean6

aProb6 = matrix(0, nrow=nrow(aObs6), ncol=ncol(aObs6))
alpha6 = 1/aMeanDist6

aProb6 = exp(-alpha*distMat6)
diag(aProb6)=0
aProb6

graphAmean6 = graph_from_adjacency_matrix(aMean6, mode='undirected')
graphAprob6 = graph_from_adjacency_matrix(aProb6, mode='undirected', weighted=T)
graphAobs6 = graph_from_adjacency_matrix(as.matrix(aObs6), mode='directed', weighted=T)

#landscape
connectance6 = edge_density(graphAmean6)
connectance6

#integral index of connectivity
AL = 63990 
nlMat6 = distances(graphAmean6)
nlMat6 [is.infinite(nlMat6)] = 1000
iicMat6 = outer(area6, area6)/(1+nlMat6)
iic6 = sum(iicMat6)/AL^2
iic6

# probability of connectivity
pStarMat6 = distances(graphAprob6, weights=E(graphAprob6)$weight)
pStarMat6 = exp(-pStarMat6)
pcNum6 = outer(area6, area6)*pStarMat6
pc6 = sum(pcNum6)/AL^2
pc6
```

### Patch 7 Removed
```{r}
nodes7 = read.csv('https://github.com/rytomey/WILD7970_Spatial/raw/main/W12_Connectivity/kite7.csv')
area7 = nodes7$area
nodes7

aObs7 = read.csv('https://github.com/rytomey/WILD7970_Spatial/raw/main/W12_Connectivity/kitemove7.csv')[,-1]
diag(aObs7) = 0
aObs7

coords7 = cbind(nodes7$XCoord, nodes7$YCoord)
distMat7 = pointDistance(coords7, lonlat=F)
distMat7 = distMat7 / 1000 # in km
distMat7

linkLoc7 = which(aObs7 > 0, arr.ind=T)
withinDisp7 = cbind(distMat7[linkLoc7], aObs7[linkLoc7])
withinDisp7 = rep(withinDisp7[,1], withinDisp7[,2])

dispExp7 = fitdist(data = withinDisp7, distr='exp', method='mle')
plot(dispExp7)
aObs7

aMeanDist7 = mean(withinDisp7)
aMeanDist7

aMean7 = matrix(0, nrow=nrow(aObs7), ncol=ncol(aObs7))
aMean7[distMat7 < aMeanDist7] = 1
diag(aMean7) = 0
aMean7

aProb7 = matrix(0, nrow=nrow(aObs7), ncol=ncol(aObs7))
alpha7 = 1/aMeanDist7

aProb7 = exp(-alpha*distMat7)
diag(aProb7)=0
aProb7

graphAmean7 = graph_from_adjacency_matrix(aMean7, mode='undirected')
graphAprob7 = graph_from_adjacency_matrix(aProb7, mode='undirected', weighted=T)
graphAobs7 = graph_from_adjacency_matrix(as.matrix(aObs7), mode='directed', weighted=T)

#landscape
connectance7 = edge_density(graphAmean7)
connectance7

#integral index of connectivity
AL = 63990 
nlMat7 = distances(graphAmean7)
nlMat7 [is.infinite(nlMat7)] = 1000
iicMat7 = outer(area7, area7)/(1+nlMat7)
iic7 = sum(iicMat7)/AL^2
iic7

# probability of connectivity
pStarMat7 = distances(graphAprob7, weights=E(graphAprob7)$weight)
pStarMat7 = exp(-pStarMat7)
pcNum7 = outer(area7, area7)*pStarMat7
pc7 = sum(pcNum7)/AL^2
pc7

#plots
plot(graphAmean, layout=coords, vertex.label=NA)
plot(graphAprob, layout=coords, edge.width=E(graphAprob)$weight, vertex.label=NA)

plot(graphAmean6, layout=coords, vertex.label=NA)
plot(graphAprob6, layout=coords, edge.width=E(graphAprob)$weight, vertex.label=NA)

plot(graphAmean7, layout=coords, vertex.label=NA)
plot(graphAprob7, layout=coords, edge.width=E(graphAprob)$weight, vertex.label=NA)

#results
connectance6
iic6
pc6

connectance7
iic7
pc7
```
Compared to the removal of patch 7, removal of patch 6 had a greater reduction of landscape connectivity in terms of the probability of connectivity as well as the integral index of connectivity. The removal of patch 7 had a greater reduction of landscape connectivity in terms of connectance which is a simple measure for unweighted graphs and may be less reliable of a metric.

$\color{red}{\text{Good. +5}}$
