---
title: "Spatial Ecology Lab 2: Making Maps"
output: html_notebook
---
Lab 2: Making Maps; Kevin Gerena

```{r}
#read location csv, load packages, convert raw data into spatial sf object
siteData = read.csv(file.choose())
require(tidyverse)
require(sf)
require(tigris)
require(geodata)
require(terra)
require(colorspace)
require(tidyterra)
require(ggnewscale)
require(cowplot)

siteSf = st_as_sf(siteData, coords=c('easting', 'northing'), crs='+proj=utm +zone=16 +datum=NAD83 +units=m +no_defs +type=crs')

plot(st_geometry(siteSf))

#add map extent boundary
studyArea = st_bbox(siteSf)
studyArea

studyArea = studyArea + c(-200, -200, 200, 200)
studyArea

studyArea = st_as_sfc(studyArea)

plot(studyArea)
```
```{r}
#add boundary file for state of Alabama
alabama = states() %>% 
  filter(NAME=='Alabama')

plot(st_geometry(alabama))

st_crs(alabama)$proj4string

alabama = st_transform(alabama, st_crs(siteSf))
```

```{r}
#plot study area in state map
plot(st_geometry(alabama))
plot(studyArea, col='red', add=T)
plot(st_geometry(siteSf), add=T)
studyArea = st_transform(studyArea, st_crs(alabama))
```

```{r}
#add elevation data, crop to study area, and plot
usaElev = elevation_30s(country='USA', path=tempdir())
alabama = st_transform(alabama, st_crs(usaElev))
alabamaElev = crop(usaElev, alabama)
studyArea = st_transform(studyArea, st_crs(alabama))
studyAreaElevation = crop(alabamaElev, studyArea)
plot(studyAreaElevation, col=sequential_hcl(13, "Terrain"))
plot(alabamaElev)
plot(st_geometry(alabama), add=T)
studyAreaElevation = project(studyAreaElevation, crs(siteSf, proj=T))
alabama = st_transform(alabama, crs(siteSf, proj=T))
```

```{r}
#plot zoomed out map
zoomedOut = ggplot()+
  geom_sf(data = alabama)+
  geom_sf(data = studyArea, fill=NA, color='red')

zoomedOut

zoomedOut = zoomedOut +
  theme_bw()+ #Gets rid of most of the colors
  theme(panel.grid=element_blank(), #Removes the gridlines
        axis.text=element_blank(), #Removes the axis text
        axis.ticks=element_blank(), #Removes the axis tick marks
        panel.border=element_blank()) #Removes the frame around the graph

zoomedOut
```

```{r}
#plot zoomed in map
zoomedIn = ggplot()+
  geom_spatraster(data = studyAreaElevation)+
  scale_fill_gradient(low='white', high='black', na.value='#99CCFF', labs(fill='Elevation'))

zoomedIn
zoomedIn = zoomedIn+
  new_scale_fill()+
  theme_bw()+
  theme(panel.grid=element_blank(),
        axis.text.x=element_text(angle=90, vjust=0.5, size=6))

zoomedIn
  zoomedIn = zoomedIn +
  geom_sf(data = siteSf, color='orange')+
  scale_shape_manual(values=19, name='Deer locations')

zoomedIn
```

```{r}
#combine maps, add legend, and finalize map
plot_grid(zoomedOut, zoomedIn)
legend = get_legend(zoomedIn)
ggdraw(legend)
supportingFig = plot_grid(zoomedOut, ggdraw(legend), ncol=1, rel_heights = c(0.5, 1))
supportingFig
plot_grid(zoomedIn, supportingFig, rel_widths=c(1, 0.5))
ggsave2('finalMap.jpg', plot=plot_grid(zoomedIn, supportingFig, rel_widths=c(1, 1)), width=5, height=4, units='in', dpi=300)
```
In this lab we evaluated the relationship between individual deer locations and elevation in Alabama. We are primarily interested in how distribute themselves across the landscape and habitat characteristics like elevation may play a role in deer resource selection. Location data was gathered at Camp Hill, AL using an unmanned aerial system to detect deer. Alabama state map gathered using the tigris package and elevation data downloaded using the usaElev function.

