---
title: "Kevin Gerena Lab 5 assignment - Point pattern analysis"
output:
  html_document:
    df_print: paged
---

```{r}
require(spatstat)
require(tidyverse)
require(terra)
```


# Challenge 1 (3 points)

Create a simulation window on a landscape stretching from 0 to 100 units in both the x and y directions. Now simulate the distribution of an organism on this landscape using a Poisson process with a lambda value of 0.01. Plot this organism's distribution. How many organisms are there on your landscape? What is the expected number of organisms that you should find on this landscape based on your simulation parameters? Why?

```{r}
set.seed(7)
Window = owin(xrange = c(0, 100), yrange = c(0, 100))
sim = rpoispp(lambda = 0.01, nsim=1, win=Window)
plot(sim)
summary(sim)
```

122 organisms in this simulation. 100 organisms are expected from the product of a window area of 10,000 and a Poisson lambda of 0.01.

# Challenge 2 (3 points)

Verify that there no linear or quadratic pattern in the distribution of your organisms using fitted Poisson process models. Show the outcome of this analysis and explain your conclusion.

```{r}
Int = ppm(sim, ~1)
Xy = ppm(sim, ~ x + y)
Xy2 = ppm(sim, ~polynom (x, y, 2))
data.frame(model = c('int', 'xy', 'xy2'),
           AIC = c(AIC(Int), AIC(Xy), AIC(Xy2)))
```
The model with the lowest AIC score and most support was the one that assumed a homogeneous point process rather than a linear or quadratic pattern.

# Challenge 3 (14 points)

I have simulated the distributions of 2 critters and provided their x-y coordinates. One of them is a non-territorial predator. The second is a critter that is preyed upon by the predator and exhibits a fear response to the predator's presence. Use the tools you learned in the lab to explore the distributions of these two critters. Identify which one is the predator and which is the prey, and provide code and a thorough justification supporting your conclusion.


```{r}
predPreyWin = owin(xrange = c(0, 100), yrange = c(0, 100))

critter1 = read.csv('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week5/species1.csv') %>% 
  mutate(org = 'critter1')

critter2 = read.csv('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week5/species2.csv') %>% 
  mutate(org = 'critter2')

critter1
critter2

cWindow = owin(xrange = c(0, 100), yrange = c(0, 100))

c1 = ppp(x=critter1$x, y=critter1$y, window = cWindow)
c2 = ppp(x=critter2$x, y=critter2$y, window = cWindow)

plot(c1)
plot(c2)
plot(density(c1, 1))
plot(density(c2, 1))
summary(c1)
summary(c2)

critters<-rbind(critter1, critter2)
critters
c3 = ppp(x=critters$x, y=critters$y, window = cWindow, marks = factor(critters$org))
plot(c3)
summary(c3)

MultiCritter2 = envelope(c3, Lcross, nsim = 99, rank = 1, i = 'critter2', j = 'critter1', simulate = expression(rlabel(pppc)))
plot(MultiCritter2, . - r ~ r, legend = FALSE)

MultiCritter1 = envelope(c3, Lcross, nsim = 99, rank = 1, i = 'critter1', j = 'critter2', simulate = expression(rlabel(pppc)))
plot(MultiCritter1, . - r ~ r, legend = FALSE)

mc=markcorr(c3)
plot(mc)

pTrans1 = pcf(c1, correction='translate')
plot(pTrans)
pEnv1 = envelope(c1, pcf, nsim=99, rank=1, correction='translate', global=F)
plot(pEnv, shade=c('hi', 'lo'), legend=F)
pTrans2 = pcf(c2, correction='translate')
plot(pTrans)
pEnv2 = envelope(c2, pcf, nsim=99, rank=1, correction='translate', global=F)
plot(pEnv, shade=c('hi', 'lo'), legend=F)

gTrans1 = Gest(c1, correction='rs')
plot(gTrans1, legend=F)
gEnv1 = envelope(c1, Gest, nsim=99, rank=1, correction='rs', global=F)
plot(gEnv1, shade=c('hi', 'lo'), legend=F)

gTrans2 = Gest(c2, correction='rs')
plot(gTrans2, legend=F)
gEnv2 = envelope(c2, Gest, nsim=99, rank=1, correction='rs', global=F)
plot(gEnv2, shade=c('hi', 'lo'), legend=F)
```
Species 1 appears to be aggregated around r=4 according to the G function while species 2 has fewer nearest neighbors around r=3 and r=6 and G(r) is generally lower than the dotted line. Species 1 appears to more closely follow a random distribution. As such, species 2 is likely the prey species avoiding species 1 as a predator.